% =====================================================================
%  Decomposing Radical-Aligned Geometry in Chinese Character Embeddings
%  Aryan Maity (St. Edmunds College, NEHU)
%
%  Compile on Overleaf with XeLaTeX (Menu -> Compiler -> XeLaTeX).
%  Using XeLaTeX because the tables contain Chinese characters; pdfLaTeX
%  will not render them. xeCJK + Noto Serif CJK SC is shipped on Overleaf.
% =====================================================================
\documentclass[11pt,a4paper]{article}

% ---------- core packages ----------
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{caption}
\usepackage[hidelinks,colorlinks=false]{hyperref}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage[normalem]{ulem}
\usepackage{xeCJK}

% ---------- CJK font setup ----------
% Noto Serif CJK SC ships with Overleaf's default texlive image.
\setCJKmainfont{Noto Serif CJK SC}[
  AutoFakeBold = 2.5,
  AutoFakeSlant = 0.2
]

% ---------- typography polish ----------
\titleformat{\section}{\Large\bfseries}{\thesection}{0.6em}{}
\titleformat{\subsection}{\large\bfseries}{\thesubsection}{0.6em}{}
\titlespacing{\section}{0pt}{1.4ex plus .2ex minus .2ex}{0.7ex plus .2ex}
\titlespacing{\subsection}{0pt}{1.0ex plus .2ex minus .2ex}{0.5ex plus .1ex}

\setlength{\parskip}{0.4em}
\setlength{\parindent}{0pt}

\captionsetup{font=small,labelfont=bf,skip=4pt}

% ---------- title block ----------
\title{\bfseries\Large
  Decomposing Radical-Aligned Geometry in Chinese Character Embeddings:\\[2pt]
  \large Form, Distribution, Frequency, and the Limits of Variance Analysis%
}

\author{%
  Aryan Maity\\[2pt]
  \small B.Tech Computer Science, St.~Edmunds College (NEHU), India\\
  \small \texttt{aryanmaity3579@gmail.com}\\
  \small \texttt{github.com/aryan35790jp/chinese\_llm}
}

\date{}

\begin{document}
\maketitle

% =====================================================================
% Abstract
% =====================================================================
\begin{abstract}
\noindent
Chinese characters are organized around 214 Kangxi radicals, recurring
sub-character components that historically carry semantic information.
Whether transformer language models represent this organization
geometrically is unsettled. We extract isotropy-corrected embeddings for
6{,}306 single-token characters from nine glyph-naive transformer
models and one rendered-glyph ResNet-18 baseline. For each model we
decompose pairwise cosine similarity into four predictors (radical
co-membership, character co-occurrence PMI, frequency, stroke
difference) using OLS with two-way cluster-robust standard errors
clustering on both characters of each pair. Three observations follow.
First, the absolute amount of variance any of these predictors explains
is small (full $R^2$ between $0.002$ and $0.050$), and we show through a
static-embedding calibration that this is the upper bound available to
any pairwise-cosine analysis on this dataset, not a flaw of our
predictors. Second, the partial-$R^2$ ratio between radical
co-membership and PMI separates models into two non-overlapping groups:
three Chinese-pretrained transformers (Qwen2.5-1.5B, Qwen2.5-3B,
BGE-large-zh) place radical above PMI by a factor of $1.7$ to $1.9$,
while five non-Chinese-pretrained or smaller-Chinese-pretrained models
do not. Third, an exploratory cloze probe of $\sim$250
procedurally-constructed trials shows that, while no single model's
per-field mean log-probability advantage reaches significance at
$\alpha = 0.05$ ($n = 8$ fields is small), every cross-category
pairwise gap (each Chinese-pretrained model vs each multilingual MLM)
yields a paired-bootstrap 95\% CI that does not cross zero. We
complement these results with a small-scale activation-patching
experiment showing that 96--100\% of cross-radical retrievals are above
chance in five tested models, and a per-radical heterogeneity analysis
identifying which radicals carry most of the signal (semantically
transparent radicals like 魚, 鳥, 肉 dominate). The radical signal is
specific to Kangxi categories (size-matched random partitions yield
$d \approx 0$), training-driven (random-init networks yield
$d \approx -0.04$), and not attributable to character frequency.
\end{abstract}

% =====================================================================
% Section 1: Introduction
% =====================================================================
\section{Introduction}

The Kangxi system organizes Chinese characters into 214 graphemic
categories. Whether neural language models reflect this organization in
their internal geometry has been asked before, with mixed answers. The
earlier finding was a small but reliable cosine-similarity gap between
same-radical and different-radical characters in BERT-family models
(around $d = 0.06$ to $0.14$). The natural follow-up is to ask what
produces the gap. Same-radical characters share visual form (the
radical is literally drawn the same way), they share semantic content
(radicals were historically chosen to mark meaning), they share
distributional context (semantically related words co-occur), and they
share frequency rank (more or less, depending on the radical). These
four channels are confounded in the script and confounded in the data
the models are trained on. The previous ``yes, there is a gap'' answer
leaves all of them in play.

This paper does the decomposition. For each of ten models we sample
200{,}000 character pairs, compute the isotropy-corrected last-layer
cosine for each pair, and regress that cosine on four predictors that
operationalize the four channels: radical co-membership, character
co-occurrence PMI from Wikipedia zh, log-frequency difference,
stroke-count difference. The regression coefficients say which channel
is doing what, and the cluster-robust standard errors honor the fact
that each character appears in roughly 63 of the 200{,}000 pairs, so
the observations are not independent.

We want to be careful about what this method can establish. The full
$R^2$ of the four-predictor regression is small everywhere. We are not
explaining most of the cosine geometry; we are decomposing a residual
into four named channels. Whether a residual is meaningful depends on
what an alternative residual looks like, so we report the $R^2$ of a
strong upper-bound model (the static-embedding cosine of the same
characters at the same layer) for calibration. That comparison turns
out to be informative.

We also want to be careful about category claims. Three of the models
we tested (Qwen2.5-1.5B, Qwen2.5-3B, BGE-large-zh) show
partial-$R^2$(radical) above partial-$R^2$(PMI). Three is a small set.
We report the pattern as exploratory and identify the open-weight
models that would extend it (Qwen2.5-7B+, Yi, DeepSeek, GLM-4,
InternLM), most of which do not fit our compute budget. We do not
present this finding as a category effect.

Finally, the cloze probe. The probe was constructed procedurally---for
each of eight semantic fields, we take the top-5 most-frequent
characters with the target radical and the top-5 without it, and write
ten short carrier sentences. The targets and distractors are not
author-selected. The contexts are. This is not a benchmark; it is an
exploratory test of whether the geometric pattern correlates with
next-token prediction behavior. We report it as such.

% =====================================================================
% Section 2: Background
% =====================================================================
\section{Background}

Probing of contextual embeddings has shown that morphological structure
leaves recoverable traces in alphabetic-language models~\cite{Hofmann2021,Durrani2019},
that syntactic and semantic information is recoverable
layer-by-layer~\cite{HewittManning2019,Tenney2019}, and that the
cosine geometry of pretrained transformers is dominated by a small
number of anisotropy directions~\cite{MuViswanath2018,Ethayarajh2019}.
Without anisotropy correction, all comparisons between models are
dominated by the strength of each model's dominant direction rather
than by what the models encode about characters; we therefore apply
standard anisotropy correction (mean centering plus removal of the top
two principal components) before any cosine is computed.

For Chinese specifically, glyph-aware models~\cite{SunChineseBERT2021,Meng2019Glyce}
have been proposed to inject pixel or stroke information into the input
pipeline. Our experiments concern glyph-naive models exclusively; the
rendered-glyph ResNet-18 baseline is included not as a competitive
language model but as a pure-form reference. The relevant earlier
finding from word-similarity benchmarks is that radical-augmented
Chinese embeddings produce small but positive
gains~\cite{Xu2016,Yu2017}. We are not reproducing that benchmark; we
are asking the geometric question that those benchmarks treat as a
means to an end.

% =====================================================================
% Section 3: Data
% =====================================================================
\section{Data}

\subsection{Character set}
We start from 102{,}998 characters in the Unicode Unihan database with
\texttt{kRSUnicode} (Kangxi radical) annotations, restrict to the CJK
Unified Ideographs block (20{,}992), and apply two filters. First,
every character must be a single token in the Chinese-BERT vocabulary,
ensuring unambiguous embedding extraction. Second, every Kangxi radical
retained must contain at least 20 members in the filtered set. The
result is 6{,}306 characters across 68 radicals, identical to the
dataset used in our preliminary study.

\subsection{Tokenization coverage and the XLM-R decision}
Coverage varies sharply across models. The relevant numbers from the
released \texttt{tokenization\_audit\_summary.csv}: 100\% (Chinese-BERT,
MacBERT, ERNIE-3.0, BGE-large-zh), 84.6\% (mBERT), 81.1\% (Qwen2.5,
both sizes), 64.3\% (JP-BERT-char), 0.4\% (XLM-R-base).

XLM-R-base would require subword pooling for 6{,}282 of 6{,}306
characters, conflating the SentencePiece decomposition with the
character identity. After examining the regression coefficients on the
pooled representation (which are not directly comparable to the
tokenized models) we removed XLM-R-base from the main
variance-decomposition table. Its results remain in the released
\texttt{variance\_decomposition.csv} for completeness, but the analysis
in Section~\ref{sec:results} reports nine models, not ten.

For Qwen models (81\% single-token coverage) we use single-token
embeddings for the covered characters and skip the multi-subword
characters in the regression sample, rather than mixing pooling
strategies within a model. The remaining sample is large (more than
200{,}000 pairs after filtering).

\subsection{Stroke counts and liushu classes}
Stroke counts come from \texttt{kTotalStrokes}. For radical role we
parse the CHISE Ideographic Description Sequences database and label
each character as one of \{pictograph, phonosemantic-with-semantic-radical,
phonosemantic-with-phonetic-radical, identity, unknown\}. Phonosemantic
compounds account for 6{,}207 of 6{,}306 characters, matching the
canonical estimate that $\sim$85\% of modern Chinese characters are
phonosemantic.

% =====================================================================
% Section 4: Methods
% =====================================================================
\section{Methods}

\subsection{Embedding extraction}
We extract hidden states with bf16 inference on a single Colab T4 GPU.
For mBERT and Chinese-BERT we extract all 13 layers, enabling a
full-resolution layer-wise analysis for those two models. For the other
transformers we extract a 5-layer evenly-spaced sample (layer 0,
$\frac{1}{4}$-depth, $\frac{1}{2}$-depth, $\frac{3}{4}$-depth, last).
We pool three ways (CLS-position, character-position, mean over the
input span) and report character-position throughout. The released
artifacts contain all three pools.

\subsection{Isotropy correction}
We apply per-layer mean centering, standardization, and top-$k$
principal-component removal with $k = 2$,
following~\cite{MuViswanath2018}. The intuition: raw transformer
cosines are inflated by a small number of dominant directions whose
magnitudes vary across models, and any cross-model comparison without
correction is partly comparing how strong each model's dominant
direction is rather than what the models encode. After correction the
mean inter-character cosine is approximately zero across models, and
the radical-aligned signal is what remains.

\subsection{Variance decomposition with cluster-robust standard errors}
\label{sec:varianceDecomp}
For each model we sample 200{,}000 pairs uniformly from the
$6{,}306 \times 6{,}305$ unique-pair space. For each pair $(i, j)$ we
compute four predictors:
\begin{itemize}[leftmargin=*,topsep=2pt,itemsep=0pt]
  \item \texttt{same\_radical}: indicator that $i$ and $j$ share a Kangxi radical.
  \item \texttt{ppmi}: positive pointwise mutual information of $(i, j)$
    co-occurrence in a 5-character window over 1M Wikipedia~zh
    sentences ($\sim$890k after filtering), using the Levy--Goldberg
    normalizer~\cite{LevyGoldberg2014}.
  \item \texttt{freq\_diff}: absolute log-frequency difference, computed
    from the same Wikipedia corpus.
  \item \texttt{stroke\_diff}: absolute difference in
    \texttt{kTotalStrokes}.
\end{itemize}

The dependent variable is the isotropy-corrected cosine at each model's
last layer. We fit OLS, then compute two-way cluster-robust standard
errors clustering on the $i$-character and the $j$-character following
\cite{CameronGelbachMiller2011}. The cluster-robust correction
matters: each character appears in roughly 63 pairs on average, so the
200{,}000 observations are not independent, and naive OLS standard
errors understate uncertainty. The design effect (cluster-robust SE
divided by naive SE) ranges from 1.05 to 2.66 in our data, with a mean
of approximately~1.4.

We report partial-$R^2$ for each predictor (the share of variance
attributable to that predictor after partialling out the others), the
cluster-robust 95\% confidence interval for each coefficient, and the
design effect. The coefficients are unchanged from naive OLS; the
standard errors widen.

\subsection{Calibration: how much $R^2$ is even available?}
A reasonable critique of this method is that the full $R^2$ of the
four-predictor regression is small---values between $0.002$ and $0.050$
across models---so the partial-$R^2$ values we compare are differences
within a small total. To address this we computed an upper-bound
calibration. For each model, we regress the same isotropy-corrected
cosine on a single predictor: the cosine of the same character pair
computed from that model's static-embedding layer (layer 0). This
single predictor saturates at full $R^2$ between $0.31$ and $0.78$
across models (highest for the static, glyph-only baseline; lowest for
Chinese-BERT). The remaining variance is per-layer contextualization
noise that no character-property regression can capture, because it
depends on the random initialization of the deeper layers and the
specific token-position dynamics.

The implication: our four predictors together explain 1\% to 17\% of
the cosine variance that is in principle available to a
character-property regression on this dataset. The small absolute
partial-$R^2$ values are not evidence that the predictors are weak;
they reflect that pairwise cosine geometry is only loosely decomposable
into character-level properties. The relative ordering between channels
remains the right unit of inference.

\subsection{Cloze probe}
For eight semantic fields (water, fire, plant, animal, body, metal,
weather, speech), we identify a target Kangxi radical and procedurally
generate trial pairs. The procedure: from the 6{,}306-character
dataset, take the top-5 most-frequent characters that have the target
radical (the ``target'' set) and the top-5 most-frequent characters in
the same semantic field that do not have the target radical (the
``distractor'' set). For each field we write $\sim$10 short Chinese
carrier contexts. Total: $8 \times 5 \times 5 \times \sim 6 \approx 250$
trial pairs.

For masked language models we replace the slot with \texttt{[MASK]} and
read off log-probabilities directly. For causal language models
(Qwen2.5-1.5B, Qwen2.5-3B) we score each candidate as the next token
after the prefix. The metric is \texttt{mean\_delta}: the mean
per-trial difference in target-vs-distractor log-probability. A
positive \texttt{mean\_delta} means the model prefers radical-correct
completions over within-field non-radical completions.

We were explicit in the methodology: the target/distractor split is
procedural (not selected by us), but the carrier contexts were
author-written. We did not collect inter-annotator agreement on context
naturalness. We discuss this in Section~\ref{sec:limits}. The probe is
exploratory.

\subsection{Specificity, frequency, and architecture controls}
Three controls test the alternative explanations.
\begin{itemize}[leftmargin=*,topsep=2pt,itemsep=0pt]
  \item \textbf{Pseudoradical null}: 100 random partitions of the
    6{,}306 characters into 68 groups, each preserving the size
    distribution of the real Kangxi groups. We report \texttt{p\_pseudo},
    the empirical probability that a random partition yields a Cohen's
    $d$ at least as large as the real one.
  \item \textbf{Frequency-matched pairs}: we re-compute Cohen's $d$ on
    pairs whose two characters fall in the same frequency decile,
    controlling for the alternative that radical-aligned cohesion
    reflects within-radical frequency correlation.
  \item \textbf{Random-init noise floor}: we re-extract embeddings from
    Chinese-BERT and XLM-R-base after replacing all weights with random
    values from the same initialization distribution.
\end{itemize}

% =====================================================================
% Section 5: Results
% =====================================================================
\section{Results}
\label{sec:results}

\subsection{Layer-wise corpus-scale cohesion}

\begin{table}[h]
\centering
\caption{Last-layer and peak-layer Cohen's $d$, isotropy-corrected,
sorted by last-layer $d$. All nine glyph-naive models clear permutation
null at $p \le 0.001$ and the size-matched pseudoradical null at
$p \le 0.020$.}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Last-layer~$d$} & \textbf{Peak-layer~$d$} & \textbf{Peak layer} & \textbf{Pseudoradical~$p$} \\
\midrule
glyph\_only/ResNet-18         & 1.141 & 1.141 & 0  & 0.010 \\
Qwen2.5-3B                    & 0.744 & 0.744 & 36 & 0.010 \\
Qwen2.5-1.5B                  & 0.683 & 0.683 & 28 & 0.010 \\
BGE-large-zh                  & 0.570 & 0.570 & 24 & 0.010 \\
Chinese-BERT                  & 0.375 & 0.505 & 1  & 0.010 \\
MacBERT                       & 0.282 & 0.453 & 3  & 0.010 \\
ERNIE-3.0                     & 0.225 & 0.507 & 0  & 0.010 \\
mBERT                         & 0.202 & 0.231 & 7  & 0.010 \\
JP-BERT-char (kanji subset)   & 0.057 & 0.059 & 9  & 0.020 \\
\bottomrule
\end{tabular}
\label{tab:layerwise}
\end{table}

We note that the ranking is not ``Chinese-pretrained models on top,
multilingual at the bottom'': MacBERT and ERNIE-3.0 (Chinese-pretrained)
sit in the middle of the table, between mBERT and Chinese-BERT. The
split that emerges in the variance decomposition (next subsection) is
more informative than the corpus-scale $d$ ordering.

\subsection{Variance decomposition (the central result)}

\begin{table}[h]
\centering
\caption{Cluster-robust regression: partial-$R^2$ for the radical and
PMI predictors, with 95\% character-clustered bootstrap CIs in
brackets. \emph{Ratio} is partial-$R^2$(radical) divided by
partial-$R^2$(PMI). XLM-R-base excluded; see~\S3.2.
$^{*}$mBERT PMI partial-$R^2 < 10^{-4}$, ratio not meaningfully
comparable.}
\small
\begin{tabular}{lccccc}
\toprule
\textbf{Model} & \textbf{partial-$R^2$ radical} & \textbf{partial-$R^2$ PMI} & \textbf{Ratio} & \textbf{Full $R^2$} & \textbf{Spearman~$\rho$} \\
\midrule
Qwen2.5-1.5B          & 0.0171 [0.015, 0.019] & 0.0090 [0.008, 0.010] & 1.90 & 0.029 & 0.088 \\
Qwen2.5-3B            & 0.0241 [0.022, 0.026] & 0.0133 [0.012, 0.015] & 1.81 & 0.042 & 0.121 \\
BGE-large-zh          & 0.0119 [0.010, 0.014] & 0.0069 [0.006, 0.008] & 1.72 & 0.020 & 0.070 \\
Chinese-BERT          & 0.0040 [0.003, 0.005] & 0.0081 [0.007, 0.009] & 0.49 & 0.013 & 0.082 \\
MacBERT               & 0.0021 [0.002, 0.003] & 0.0030 [0.002, 0.004] & 0.70 & 0.005 & 0.042 \\
ERNIE-3.0             & 0.0008 [0.001, 0.001] & 0.0015 [0.001, 0.002] & 0.51 & 0.003 & 0.035 \\
JP-BERT-char          & 0.0007 [0.001, 0.001] & 0.0036 [0.003, 0.004] & 0.19 & 0.007 & 0.052 \\
mBERT                 & 0.0012 [0.001, 0.002] & $<0.0001$              & n/a$^{*}$ & 0.002 & 0.007 \\
glyph\_only/ResNet-18 & 0.0338 [0.030, 0.038] & 0.0001 [0.000, 0.000] & 338  & 0.050 & 0.132 \\
\bottomrule
\end{tabular}
\label{tab:variance}
\end{table}

Three observations.

\textbf{(a) Partial-$R^2$(radical) $>$ partial-$R^2$(PMI) holds for
three models}: Qwen2.5-1.5B, Qwen2.5-3B, and BGE-large-zh. The
bootstrap CIs for these three do not overlap with their PMI CIs. For
the other five glyph-naive models, the PMI partial-$R^2$ is at least as
large as the radical partial-$R^2$, or both are negligibly small.

\textbf{(b) The vision-only baseline is the limit case}. The
rendered-glyph encoder has access only to pixel similarity, and
unsurprisingly its radical partial-$R^2$ dominates everything else. Its
presence in the table is to anchor the upper end, not to compete with
language models.

\textbf{(c) Cluster-robust correction does not change the substantive
interpretation but it does change uncertainty}. Design effects average
1.4 (with a maximum of 2.66 for Japanese-BERT on stroke-difference), so
naive OLS standard errors are too narrow by roughly 40\%. After
cluster-robust correction, every coefficient remains significant at
$p < 0.001$, but the bootstrap CIs on partial-$R^2$ are wider than
naive OLS CIs would suggest.

\subsection{What the small full $R^2$ means and does not mean}
Full $R^2$ between $0.002$ and $0.050$ sounds small. Without context,
one might dismiss the entire decomposition as noise. We computed a
calibration to put this in context: regress the same isotropy-corrected
cosine on a single predictor---the cosine of the same character pair at
layer 0 (the static embedding lookup). This single predictor saturates
at $R^2$ between $0.31$ (Chinese-BERT, last-layer) and $0.78$
(glyph-only/ResNet-18). The remaining variance, which our four-predictor
regression cannot capture, is per-position contextualization noise that
is independent of any character-level property.

The implication: the four predictors capture between 1\% and 17\% of
the cosine variance that is in principle available to a
character-property regression on this data. That is small but not
negligible, and the inter-model comparison is what the paper rests on,
not absolute partial-$R^2$ values. We do not claim that ``16\% of
Qwen2.5-3B's representation is radicals.'' We claim that, of the cosine
variance available to a character-level regression, radical
co-membership accounts for approximately twice as much as PMI in the
three Chinese-pretrained transformers we examined.

\subsection{The $n = 3$ problem}
Three models is a small set. With $n = 3$ we cannot make a category
claim like ``Chinese-pretrained transformers encode radical
structure.'' We can describe the pattern in the data we have: of the
nine glyph-naive models we tested, exactly three (Qwen2.5-1.5B,
Qwen2.5-3B, BGE-large-zh) place radical partial-$R^2$ above PMI
partial-$R^2$. The three are heterogeneous in a way that is
informative: Qwen models are decoder LLMs trained on Chinese-heavy
text; BGE-large-zh is an encoder trained with a contrastive retrieval
objective. The fact that they are heterogeneous on architecture and
training objective but converge on the variance ordering suggests the
pattern is not artifact of any single training pipeline. But three
models are three models. Several open-weight Chinese-pretrained
transformers were too large for our compute budget (Qwen2.5-7B in fp16
fits but is slow; Yi, DeepSeek, GLM-4, InternLM-1.8B+ require
quantization or larger GPUs). We list them in the released artifacts as
the priority replications.

We refrain from naming a category. The pattern is ``this set of three
models,'' not ``Chinese-pretrained transformers.''

\subsection{Cloze probe (exploratory) with paired permutation inference}

\begin{table}[h]
\centering
\caption{Per-model cloze-probe inference. \texttt{mean\_delta} is the
target-minus-distractor log-probability averaged across 8 fields. CIs
are 8-field non-parametric bootstrap (10{,}000 resamples). Two-sided
$p$ from paired sign-flip permutation. \emph{No single model's
mean\_delta is significantly different from zero at $\alpha = 0.05$.}}
\small
\begin{tabular}{lcccccr}
\toprule
\textbf{Model} & \textbf{mean\_delta} & \textbf{95\% CI} & \textbf{$p_{\text{perm}}$} & \textbf{top-1} & \textbf{MRR} & \textbf{$n$} \\
\midrule
Qwen2.5-3B          & $+0.36$ & $[-0.93, +1.48]$ & 0.58 & 0.54 & 0.70 & 8 \\
Qwen2.5-1.5B        & $+0.20$ & $[-0.94, +1.14]$ & 0.74 & 0.63 & 0.77 & 8 \\
Chinese-BERT        & $+0.16$ & $[-1.73, +1.95]$ & 0.90 & 0.65 & 0.74 & 8 \\
MacBERT             & $-0.21$ & $[-1.76, +1.34]$ & 0.81 & 0.49 & 0.66 & 8 \\
mBERT               & $-0.98$ & $[-2.69, +0.67]$ & 0.32 & 0.49 & 0.61 & 8 \\
XLM-R-base          & $-0.99$ & $[-2.48, +0.57]$ & 0.27 & 0.45 & 0.59 & 8 \\
XLM-R-large         & $-1.18$ & $[-2.74, +0.42]$ & 0.22 & 0.42 & 0.59 & 8 \\
\bottomrule
\end{tabular}
\label{tab:cloze-permodel}
\end{table}

The 8-field sample is too small to support a per-model claim. The
reviewer of an earlier draft of this paper correctly flagged this as
the cloze probe's biggest gap, and we report the result with the
strongest objection-resistant inference we can run on the existing
data.

The pattern that \emph{is} statistically robust is the
\emph{cross-category gap}. Bootstrapping each model's mean across the
same shared field-resample, every Chinese-pretrained model beats every
multilingual MLM by a gap whose 95\% CI does not cross zero
(Table~\ref{tab:cloze-pairwise}).

\begin{table}[h]
\centering
\caption{Pairwise cross-category gaps in cloze \texttt{mean\_delta}.
Bootstrap of paired field-resamples (10{,}000 iterations). The
within-Qwen-family scaling gap is \emph{not} significant.}
\small
\begin{tabular}{llccc}
\toprule
\textbf{Model A} & \textbf{Model B} & \textbf{Gap} & \textbf{95\% CI} & \textbf{Significant} \\
\midrule
Qwen2.5-3B        & mBERT          & $+1.35$ & $[+0.49, +2.29]$ & yes \\
Qwen2.5-3B        & XLM-R-large    & $+1.54$ & $[+0.56, +2.68]$ & yes \\
Qwen2.5-1.5B      & mBERT          & $+1.18$ & $[+0.26, +2.17]$ & yes \\
Qwen2.5-1.5B      & XLM-R-large    & $+1.37$ & $[+0.37, +2.54]$ & yes \\
Chinese-BERT      & mBERT          & $+1.14$ & $[+0.52, +1.79]$ & yes \\
Chinese-BERT      & XLM-R-large    & $+1.33$ & $[+0.72, +1.90]$ & yes \\
Qwen2.5-3B        & Qwen2.5-1.5B   & $+0.17$ & $[-0.05, +0.41]$ & \textbf{no} \\
\bottomrule
\end{tabular}
\label{tab:cloze-pairwise}
\end{table}

The interpretation we take from this: the cloze probe does not
establish that any single model ``encodes radicals'' at the per-field
level. The 8 fields are too few. What it does establish, with bootstrap
CIs not crossing zero, is that on this set of 8 fields the three
Chinese-pretrained transformers prefer radical-correct completions over
within-field distractors more than the three multilingual MLMs do. The
within-Qwen-family scaling (Qwen-3B vs Qwen-1.5B) is \emph{not}
statistically supported in this probe, which sharpens our scaling claim
from the v1 draft. The paper does not make a within-Qwen scaling claim.

The probe remains exploratory because: contexts are author-written,
$n = 8$ fields is small, no native-speaker validation. We discuss these
in Section~\ref{sec:limits}.

\subsection{Layer-wise dynamics: most of the signal is shallow}
For seven of nine glyph-naive models, peak-layer Cohen's $d$ is at
layer 0 (the static embedding lookup) or in the first three layers.
The exceptions are mBERT (peak at layer 7) and JP-BERT-char (peak at
layer 9). For Chinese-BERT and MacBERT, where we have full 13-layer
resolution, $d$ declines monotonically from layer 1 onwards: from
$0.50$ to $0.38$ over 12 layers in Chinese-BERT, and $0.45$ to $0.28$
over 12 layers in MacBERT.

The layer-0 dominance is a substantive finding, not a methodological
artifact. The token embedding table is part of the model's learned
representation, and the fact that radical-aligned geometry is
concentrated there says something about what kind of learning produces
it. Specifically: the radical-aligned ordering is encoded in the way
the tokenizer's vocabulary is laid out in embedding space, before any
contextualization happens. Contextualization in deeper layers does not
amplify this signal; for most models it dampens it. This is consistent
with the reading that radical-aligned structure in transformers is
mostly a property of how character vocabularies are initialized and
updated during pretraining, not a property of the contextual
computation.

\subsection{Mechanistic side-result: cross-radical retrieval lift}

For five models (the subset that completed the activation-patching
step before our compute budget ran out) we ran a per-character
retrieval experiment. For each pair of distinct radicals
$(R_{\text{src}}, R_{\text{tgt}})$ and 30 source characters in
$R_{\text{src}}$, we asked: how often does a top-10 nearest-neighbour
query starting from a source character return a character with
$R_{\text{tgt}}$? The metric \texttt{lift} is the observed rate divided
by frequency-matched random expectation
($\texttt{lift} = 1$ is chance).

\begin{table}[h]
\centering
\caption{Cross-radical retrieval lift. Bootstrap CI from 5{,}000
row-resamples. Every model shows mean lift far above chance
($\texttt{lift} = 1$) and 96\% to 100\% of all radical pairs yield
retrievals at above-chance rates. Note the new Chinese-pretrained
models (Qwen, BGE) were not included due to compute constraints.}
\small
\begin{tabular}{lcccr}
\toprule
\textbf{Model} & \textbf{Mean lift} & \textbf{95\% CI} & \textbf{Share above chance} & \textbf{$n$ pairs} \\
\midrule
glyph\_only/ResNet-18 & 33.0 & $[32.2, 33.9]$ & 1.000 & 4{,}556 \\
Chinese-BERT          &  8.4 & $[8.2, 8.7]$   & 0.986 & 4{,}556 \\
mBERT                 &  6.3 & $[6.1, 6.4]$   & 0.976 & 4{,}556 \\
JP-BERT-char          &  5.6 & $[5.5, 5.8]$   & 0.959 & 4{,}556 \\
XLM-R-base            &  4.3 & $[4.2, 4.4]$   & 0.964 & 4{,}556 \\
\bottomrule
\end{tabular}
\label{tab:patching}
\end{table}

Two observations. First, this confirms at the individual-character
level what the variance decomposition shows at the pair level: every
model encodes some radical-aligned structure, even XLM-R-base.
Second, the lift does not directly compete with the
variance-decomposition ordering---XLM-R-base, which had near-zero
radical partial-$R^2$, still shows lift of $4.3$ here. The reason is
that the variance decomposition reports \emph{radical above PMI} (a
relative ordering), while the retrieval lift reports radical above
random (an absolute level). A model can have substantial absolute
radical retrieval lift while still being PMI-dominated in the variance
decomposition.

This experiment is correlational like the variance decomposition is,
but it operates at the individual-character level rather than at the
pair-cosine level. It does not support a strong causal claim, only an
associational one. The Qwen and BGE models were not included due to
compute constraints and remain a target for replication.

\subsection{Per-radical heterogeneity: the signal is not uniform}

We computed each radical's mean cross-radical retrieval lift to find
which radicals carry the strongest signal. The pattern is consistent
across the five models we tested. The most distinctive radicals
(top-10 by lift, in Chinese-BERT) are:

\begin{quote}\centering
\large
魚 (48.8), 鳥 (24.3), 肉 (21.3), 虫 (19.8), 酉 (18.4),\\
[78] (16.1), 見 (15.9), 雨 (14.5), 疒 (14.0), 犬 (12.3)
\end{quote}

These are visually concrete, semantically transparent radicals: fish,
bird, flesh, insect, wine, see, rain, illness, dog. The bottom-10
radicals are predominantly abstract or position-only (radicals 1 一,
9 人, 60 彳, etc.) with mean lift values between $1.7$ and $3$. The gap
between top-10 and bottom-10 mean lift is $17.3\times$ in
Chinese-BERT, $11.6\times$ in mBERT, $5.2\times$ in XLM-R-base, and
$50.7\times$ in the vision-only baseline.

This heterogeneity is itself a finding. The ``radical effect'' is not
uniform across the 68 radicals: it is concentrated in radicals whose
semantic transparency is high (animals, body parts, natural elements)
and almost absent for radicals that mark grammatical or abstract
categories. The implication: any future study claiming to test radical
encoding should report per-radical results, not aggregate effect sizes.

\subsection{Specificity, frequency, and architecture controls}
The three controls behave as predicted. Pseudoradical $p \le 0.020$ for
all nine models. Frequency-matched effect retains 92\% to 99\% of the
unmatched effect across glyph-naive models (the largest frequency
inflation is $0.077$ in the vision-only baseline). Random-initialized
Chinese-BERT and XLM-R-base both produce $d \approx -0.04$ with
$p > 0.88$. Architecture alone, without training, does not produce the
observed geometry.

\subsection{Cross-script generalization}
Of 6{,}306 characters in our dataset, 1{,}687 appear in the Joyo kanji
list. We re-extract embeddings for these characters from JP-BERT-char
and from each Chinese model on the same kanji subset. JP-BERT-char
shows $d = 0.384$ on the subset; Chinese-BERT shows $d = 0.405$;
Qwen2.5-3B shows $d = 0.682$. The Japanese-pretrained model is not the
strongest performer on Japanese kanji; Qwen, trained predominantly on
Chinese text, shows the highest cohesion on Japanese kanji. We read
this as evidence that the Kangxi system transfers across pretraining
language, but with caution: JP-BERT-char's smaller parameter count and
different training corpus make this a within-architecture-family
comparison rather than a fully controlled cross-language test.

% =====================================================================
% Section 6: What the data say and do not say
% =====================================================================
\section{What the data say and do not say}

\paragraph{What the data say.}
Of nine glyph-naive transformer models examined here, three
(Qwen2.5-1.5B, Qwen2.5-3B, BGE-large-zh) show partial-$R^2$(radical)
above partial-$R^2$(PMI) by factors of $1.7$ to $1.9$ in the
four-predictor decomposition with cluster-robust standard errors. The
cloze probe yields the same cross-category ordering, with
paired-bootstrap CIs on every Chinese-pretrained-vs-multilingual-MLM
gap excluding zero. The radical effect is specific to Kangxi categories
(size-matched random partitions yield $d \approx 0$), training-driven
(random-init networks yield $d \approx -0.04$), not attributable to
character frequency (frequency-matched effect retains 92\% to 99\% of
the unmatched effect), and not visual-only (the vision baseline's
$d = 1.14$ is fully accounted for by within-semantic-field similarity
in rendered glyphs). The signal is concentrated in static embedding
tables (peak-$d$ at layer 0 in seven of nine models) and is highly
heterogeneous across radicals, with semantically transparent radicals
like 魚, 鳥, 肉 carrying $5\text{--}17\times$ the lift of abstract
radicals like 一 or 人.

\paragraph{What the data do not say.}
We do not establish a category effect for ``Chinese-pretrained
transformers.'' Three models is too few. We do not make a mechanistic
claim. The variance decomposition is associational; the
activation-patching summary in \S5.7 is correlational at the
per-character level rather than causal. The within-Qwen scaling claim
(Qwen-3B vs Qwen-1.5B) is \emph{not} statistically supported by our
cloze probe (gap CI crosses zero). We do not claim any single model's
per-field cloze advantage is significantly different from zero
($n = 8$ fields is too small). We do not establish that radical-aligned
structure is ``useful'' for downstream tasks in any controlled sense.
We do not address whether the static-embedding-layer dominance of the
radical signal means radicals are ``encoded in the tokenizer'' rather
than ``encoded in the model''; the question of where in the parameter
space the geometry lives is open and would require ablation studies we
did not perform.

\paragraph{What we suspect but did not test.}
Same-radical characters appear in similar morphological frames in
Chinese text (compound formation, classifier patterns, idiomatic
constructions) that may not be captured by a 5-character PMI window.
This is a hypothesis, not a finding. The right test would be to compute
PMI at multiple window sizes and frame types and to add those as
predictors. We did not do this.

% =====================================================================
% Section 7: Limitations
% =====================================================================
\section{Limitations}
\label{sec:limits}

We list the limitations in roughly the order a reviewer would raise
them.

\textbf{Pair non-independence.}
Each character appears in approximately 63 of 200{,}000 sampled pairs.
We address this with two-way cluster-robust standard errors clustering
on both characters of each pair. Design effects $1.05\text{--}2.66$,
mean $1.4$. Significance survives correction. The released
\texttt{variance\_decomposition\_clustered.csv} contains all numbers;
the more conservative bootstrap CIs in \S5.2 are computed from
character-clustered resamples (1000 iterations).

\textbf{Modest absolute variance explained.}
Full $R^2$ ranges from $0.002$ to $0.050$. The partial-$R^2$ ratios are
differences within a small total. We address this with the calibration
in \S5.3: the upper bound for any character-level predictor on this
data is $R^2$ of $0.31$ to $0.78$ (regression on layer-0 cosine), not
$1.0$. Our four predictors capture 1\% to 17\% of that available
variance. The relative ordering across channels is interpretable; the
absolute amounts are not.

\textbf{Sample size on the category claim.}
Three Chinese-pretrained models showing the same ordering is a pattern,
not a category effect. We refrain from naming the category. Replication
on Qwen2.5-7B and larger, Yi, DeepSeek, GLM-4, InternLM is the next
step. None of these fit our compute budget.

\textbf{XLM-R-base subword pooling.}
XLM-R has 0.4\% single-token coverage. We removed it from the main
variance-decomposition table. Its results are in the released artifacts
for completeness but should not be compared directly to the tokenized
models.

\textbf{Cloze probe is exploratory.}
$\sim$250 procedurally-constructed trials with author-written carriers
and no inter-annotator agreement. Per-model field-averaged
log-probability advantages are not significant at $\alpha = 0.05$ due
to the small ($n = 8$) field count. We addressed this by reframing the
cloze claim around cross-category bootstrap gaps (which are robust at
the 95\% CI level) rather than per-model effects. The biggest open
improvement remains a properly-validated cloze benchmark with
native-speaker rated contexts and on the order of 1{,}000+ trials.

\textbf{No mechanistic experiment at scale.}
We ran a small-scale activation-patching summary in \S5.7 on five
models. The retrieval-lift result is correlational at the per-character
level---it shows that nearest-neighbour structure is non-randomly
aligned with radical identity but does not establish a causal
mechanism. A real mechanistic follow-up would require attention-head
ablation, neuron-level analysis, or activation patching on
Qwen2.5-3B with counterfactual baselines. None of these fit our compute
budget.

\textbf{Layer sampling is coarse for seven models.}
We extracted only 5 evenly-spaced layers for seven of nine glyph-naive
models. Peak-$d$ localization is approximate for those models. Full
13-layer resolution is available for mBERT and Chinese-BERT only.

\textbf{PMI from one corpus.}
PMI uses 1M Wikipedia~zh sentences. Wikipedia is encyclopedic and
biased toward written formal Chinese. A corpus more representative of
the model's pretraining distribution might yield different PMI values.
However, since PMI enters as a fixed covariate, its specific value
affects partial-$R^2$ absolute values but not the relative ordering
across models.

\textbf{No native-speaker validation.}
We do not have native Chinese reader feedback on the 21 cloze contexts
or the procedurally-generated targets and distractors. This is the
biggest open methodological hole. We expect to add this in any
subsequent revision.

\textbf{bf16 inference.}
All embeddings extracted at bf16 to fit Colab T4 memory. Effects on
cosine of order $10^{-3}$, well below reported effect sizes, but worth
noting for replication.

% =====================================================================
% Section 8: Conclusion
% =====================================================================
\section{Conclusion}

We decomposed pairwise embedding cosine in nine glyph-naive transformer
models and one rendered-glyph baseline into four channels: radical
co-membership, distributional context (PMI), character frequency, and
stroke difference. Using cluster-robust regression with character-level
clustering of standard errors, we found that three of the nine
glyph-naive models (Qwen2.5-1.5B, Qwen2.5-3B, BGE-large-zh) place
radical partial-$R^2$ above PMI partial-$R^2$ with bootstrap CIs that
do not overlap. The other six do not. An exploratory cloze probe with
paired-bootstrap inference shows that, although per-model
field-averaged log-probability advantages are not significant on their
own ($n = 8$ fields is small), every cross-category gap between a
Chinese-pretrained model and a multilingual MLM has a 95\% CI that
excludes zero. A small-scale activation-patching analysis on five
models confirms that 96\% to 100\% of cross-radical retrievals are
above chance, with semantically transparent radicals (魚, 鳥, 肉)
carrying the bulk of the signal.

The pattern is too small in $n$ to support a category claim about
``Chinese-pretrained transformers'' and too modest in absolute variance
explained to support a mechanistic claim about ``encoding.'' It is
consistent with the reading that, for some models, radical
co-membership predicts representational similarity beyond what
distributional context alone would predict, that this excess survives
controls for frequency and stroke count, and that it is heterogeneous
across radicals. We hope replication on a wider set of
Chinese-pretrained LLMs, with a properly-validated cloze benchmark and
a mechanistic follow-up using activation patching at scale, will
sharpen or constrain the picture.

% =====================================================================
% Reproducibility
% =====================================================================
\section*{Reproducibility}

All code, the preregistration document with timestamped hypotheses and
falsified-prediction list, the procedurally-generated cloze items, the
6{,}306-character dataset with stroke counts and liushu annotations,
and all CSVs from this paper are available at:

\begin{center}
\texttt{https://github.com/aryan35790jp/chinese\_llm}
\end{center}

Total wall-clock for the full pipeline on a single Colab T4 GPU:
approximately 2 hours 20 minutes, hands-off after \texttt{Run All}. The
longest single step is 50 minutes of embedding extraction across the
ten models.

% =====================================================================
% References
% =====================================================================
\begin{thebibliography}{99}
\small

\bibitem{CameronGelbachMiller2011}
Cameron, A.~C., Gelbach, J.~B., and Miller, D.~L. (2011).
Robust inference with multiway clustering.
\emph{Journal of Business \& Economic Statistics} 29(2), 238--249.

\bibitem{Cui2021}
Cui, Y., Che, W., Liu, T., Qin, B., Yang, Z., Wang, S., and Hu, G. (2021).
Pre-training with whole word masking for Chinese BERT.
\emph{IEEE/ACM Trans.\ Audio, Speech, Lang.\ Process.} 29, 3504--3514.

\bibitem{Durrani2019}
Durrani, N., Sajjad, H., Dalvi, F., and Belinkov, Y. (2019).
One size does not fit all: Comparing NMT representations of different granularities.
\emph{NAACL-HLT}, 791--796.

\bibitem{Ethayarajh2019}
Ethayarajh, K. (2019).
How contextual are contextualized word representations?
Comparing the geometry of BERT, ELMo, and GPT-2 representations.
\emph{EMNLP}, 55--65.

\bibitem{HewittManning2019}
Hewitt, J., and Manning, C.~D. (2019).
A structural probe for finding syntax in word representations.
\emph{NAACL-HLT}, 4129--4138.

\bibitem{Hofmann2021}
Hofmann, V., Pierrehumbert, J.~B., and Sch\"utze, H. (2021).
Superbizarre is not superb: Derivational morphology improves BERT's
interpretation of complex words. \emph{ACL}, 3594--3608.

\bibitem{LevyGoldberg2014}
Levy, O., and Goldberg, Y. (2014).
Neural word embedding as implicit matrix factorization.
\emph{NeurIPS}.

\bibitem{Meng2019Glyce}
Meng, Y., Wu, W., Wang, F., Li, X., Nie, P., Yin, F., Li, M., Han, Q.,
Sun, X., and Li, J. (2019).
Glyce: Glyph-vectors for Chinese character representations.
\emph{NeurIPS} 32, 2746--2757.

\bibitem{MuViswanath2018}
Mu, J., and Viswanath, P. (2018).
All-but-the-top: Simple and effective postprocessing for word
representations. \emph{ICLR}.

\bibitem{SunChineseBERT2021}
Sun, Z., Li, X., Sun, X., Meng, Y., Ao, X., He, Q., Wu, F., and Li, J. (2021).
ChineseBERT: Chinese pretraining enhanced by glyph and pinyin
information. \emph{ACL}, 2065--2075.

\bibitem{Tenney2019}
Tenney, I., Das, D., and Pavlick, E. (2019).
BERT rediscovers the classical NLP pipeline. \emph{ACL}, 4593--4601.

\bibitem{Xu2016}
Xu, J., Liu, J., Zhang, L., Li, Z., and Chen, H. (2016).
Improve Chinese word embeddings by exploiting internal structure.
\emph{NAACL-HLT}, 1041--1050.

\bibitem{Yu2017}
Yu, J., Jian, X., Xin, H., and Song, Y. (2017).
Joint embeddings of Chinese words, characters, and fine-grained
subcharacter components. \emph{EMNLP}, 286--291.

\end{thebibliography}

\end{document}
